---
title: "Covid19 vaccine study Final 2025"
author: "Anicia Friend"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter (FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
druguser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

```{r}
dd2 <- male %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The graph shows that vaccinated males were much less likely to get COVID-19 than those who took the placebo. Fewer than 30% of vaccinated males got sick, compared to over 60% of those who received the placebo.

### Numerical

```{r}
table1 <- xtabs(~infected + treatment, data=male)
rowPerc(table1)
colPerc(table1)
```

The numbers show that males who took the placebo were significantly more likely to get COVID-19 73.80% compared to just 26.20% of vaccinated males. Among those who didn’t get sick, the split was nearly even: 50.01% had taken the placebo and 49.99% the vaccine. The likelihood of getting COVID-19 for placebo vs. vaccinated males was also shown as a 3:1 ratio, reinforcing that placebo recipients were much more at risk.


Based on this data, it’s recommended that males get the COVID-19 vaccine, as it significantly lowers the risk of infection.


### Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
malefisher <- fisher.test(table1)
malefisher
```

Three tests were conducted to determine if death from COVID-19 is independent of a person’s sex. All tests produced extremely small p-values (2.2e-16, 0, and 2.2e-16), suggesting a strong statistical link between sex and COVID-19 mortality. The low p-values also support the vaccine's effectiveness in males. Additionally, the odds ratio shows how much more likely unvaccinated males (placebo group) were to get COVID-19 compared to vaccinated males.


### Confidence Interval

```{r}
effect <- malefisher$p.value/(malefisher$p.value +1)
effectivenessmale = malefisher$estimate/(malefisher$estimate +1)*100
print(effectivenessmale)
```

```{r}
low=(malefisher$conf.int[1]/(malefisher$conf.int[1] +1))*100
hi=(malefisher$conf.int[2]/(malefisher$conf.int[2] +1))*100
options(digits=2)
```


The data above shows the COVID-19 vaccine is effective for males, as its effectiveness is above 50%. The confidence interval ranges from 68.85% to 78.32%, confirming its reliability. The odds ratio of 73.8 further supports how much more likely unvaccinated males were to get infected compared to those who were vaccinated.


## Females



### Graphical Results

```{r}
dd2 <- female %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```


The graph shows that vaccinated females were less likely to get COVID-19 than those who took the placebo, with under 40% of vaccinated females getting sick compared to over 60% in the placebo group. However, a notable number of vaccinated females still contracted the virus.



### Numerical

```{r}
table1 <- xtabs(~infected + treatment, data=female)
rowPerc(table1)
colPerc(table1)
```


The numbers show that females who took the placebo were 64.93% more likely to get COVID-19, compared to 35.07% of vaccinated females. Among those who didn’t get sick, the split was nearly even: 49.94% had taken the placebo and 50.06% the vaccine. The infection ratio between placebo and vaccinated females was 3:1, meaning placebo recipients were three times more likely to get sick. While a significant number of vaccinated females still contracted COVID-19, the data supports recommending the vaccine, as it clearly reduces the risk of infection.



### Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
femalefisher <- fisher.test(table1)
femalefisher
```

Three tests were conducted to examine whether COVID-19 death is independent of sex. The p-values 1.914e-08, 0, and 1.246e-08 are all extremely small, indicating a strong statistical relationship between sex and COVID-19 mortality. These low values also suggest the vaccine was effective in females. The odds ratio of 1.86 shows that females who took the placebo were nearly twice as likely to get COVID-19 compared to those who were vaccinated.




### Confidence Interval


```{r}
effect <- femalefisher$p.value/(femalefisher$p.value +1)
effectivenessfemale = femalefisher$estimate/(femalefisher$estimate +1)*100
print(effectivenessfemale)
```

```{r}
low=(femalefisher$conf.int[1]/(femalefisher$conf.int[1] +1))*100
hi=(femalefisher$conf.int[2]/(femalefisher$conf.int[2] +1))*100
options(digits=2)
```


The data above shows that the COVID-19 vaccine is effective for females, as the effectiveness exceeds 50%. Based on the confidence interval provided, we can conclude that the vaccine significantly reduces the risk of infection in females.


The odds ratio, which reflects the vaccine’s effectiveness, is about 65 for females. The confidence interval for vaccine effectiveness in females ranges from 59.80% to 69.92%, confirming its strong protective effect.


## LGBTQ

### Graphical Results

```{r}
dd2 <- lgbtq %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The graph shows that LGBTQ individuals who took the vaccine were much less likely to get COVID-19 compared to those who took the placebo. About 20% of vaccinated LGBTQ individuals got sick, while nearly 80% of those in the placebo group did. This suggests the vaccine is highly effective for this group.


### Numerical Results

```{r}
table1 <- xtabs(~infected + treatment, data=lgbtq)
rowPerc(table1)
colPerc(table1)
```

The data shows that LGBTQ individuals who took the placebo were 80% more likely to get COVID-19, compared to 20% of those who were vaccinated. Among those who didn’t get sick, 48% had taken the placebo and 52% the vaccine. The infection ratio between placebo and vaccinated LGBTQ individuals was roughly 3:1, meaning those who took the placebo were much more likely to contract COVID-19 than those vaccinated.


This data clearly supports recommending the COVID-19 vaccine for LGBTQ individuals, as it significantly lowers their risk of infection.


### Inferential Results

```{r}
chisq.test(table1)
chisqtestGC(table1)
lgbtqfisher <- fisher.test(table1)
lgbtqfisher
```


Three tests were conducted to determine if COVID-19 death is independent of sexual orientation. The p-values 2e-06, 0, and 7e-07 are all extremely low, indicating a strong statistical link between sexual orientation and COVID-19 mortality. The very small p-values also suggest the vaccine’s effectiveness in LGBTQ individuals. The odds ratio of 4.3 shows that LGBTQ individuals who took the placebo were over four times more likely to get COVID-19 compared to those vaccinated.



### Confidence Interval

```{r}
effect <- lgbtqfisher$p.value/(lgbtqfisher$p.value +1)
effectivenesslgbtq = lgbtqfisher$estimate/(lgbtqfisher$estimate +1)*100
print(effectivenesslgbtq)
```

```{r}
low=(lgbtqfisher$conf.int[1]/(lgbtqfisher$conf.int[1] +1))*100
hi=(lgbtqfisher$conf.int[2]/(lgbtqfisher$conf.int[2] +1))*100
options(digits=2)
```

The data above shows that the COVID-19 vaccine is effective for LGBTQ individuals, with effectiveness well above 50%. Based on the confidence interval of 69% to 90%, we can confidently conclude the vaccine significantly reduces their risk of getting COVID-19. The odds ratio of 81 further highlights how much more likely unvaccinated LGBTQ individuals were to contract the virus compared to those vaccinated.



## Druggies








# Overall Results and Conclusions